Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-08 paper-new/concrete-relational-model.tex
[paper] small wording changes

Importantly, we note that as with the original ordering $\semltbad$ defined at
the beginning of this section, the relation just defined is not transitive
(again as a result of the above no-go theorem). It may therefore seem that we
have not solved the original issue we faced. In a sense, this is true. The model
of gradual typing that we end up with will not support compositional reasoning.
However, by decomposing the denotation of term precision in the above manner, we
can employ compositional reasoning provided we work only with the
\emph{lock-step error ordering}. Thus, the tentative plan going forward will be
to carry out the proofs compositionally using the lock-step ordering. We will
then apply the closure under weak bisimilarity in the denotation of term
precision given above to handle the aspects involving stepping.

@@

We have seen how we can recover transitive reasoning by working with the
lock-step error ordering. However, we cannot completely isolate the reasoning
about steps from the reasoning about the ordering.
% Unfortunately, by defining the partial order on the guarded lift monad to be
% the lock-step error ordering, we introduce a new issue.
In particular, the cast rules required for proving graduality no longer hold if
interpreted using the lock-step ordering. The source of the problem is that the
downcast from the dynamic type to a function takes a step, i.e., involves a
$\theta$. Consider the DnL rule shown below:
%
\begin{mathpar}
  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
             {\dnc{c}{M} \ltdyn N : c}
\end{mathpar}
%
Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
definition of the semantics of a downcast given in Section
\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:

>>>>

Importantly, we note that as with the original ordering $\semltbad$ defined at
the beginning of this section, the relation just defined is not transitive
(again as a result of the above no-go theorem). It may therefore seem that we
have not solved the original issue we faced. In a sense, this is true. The model
of gradual typing that we end up with will not support \emph{extensional}
compositional reasoning. However, by decomposing the denotation of term
precision in the above manner, we can employ \emph{intensional} compositional
reasoning by working with the lock-step error ordering. Thus, the tentative plan
going forward will be to carry out the proofs compositionally using the
lock-step ordering. We will then apply the closure under weak bisimilarity in
the denotation of term precision given above to handle the aspects involving
stepping.

@@

We have seen how we can recover transitive reasoning by working with the
lock-step error ordering. However, it turns out that we cannot completely
isolate the reasoning about steps from the reasoning about the ordering.
% Unfortunately, by defining the partial order on the guarded lift monad to be
% the lock-step error ordering, we introduce a new issue.
In particular, the cast rules required for proving graduality no longer hold if
interpreted using the lock-step ordering. The source of the problem is that the
downcast from the dynamic type to a function takes a step, i.e., involves a
$\theta$. Consider the DnL rule shown below:
%
\begin{mathpar}
  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
             {\dnc{c}{M} \ltdyn N : c}
\end{mathpar}
%
Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
definition of the semantics of a downcast given in Section
\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:
